{\steve}: A hierarchical Bayesian model for Supernova Cosmology

### 2019-01-28 papers/methods/paper.tex
Updating tikz
Alex's suggestions

At the most fundamental level, a supernova cosmology analysis is a mapping from an underlying population onto an observed population, where cosmological parameters are encoded directly in the mapping function. The difficulty arises both in adequately describing the biases in the mapping function, and in adding sufficient, physically motivated flexibility in both the observed and underlying populations whilst not adding \textit{too} much flexibility, such that model fitting becomes pathological due to increasing parameter degeneracies within the model. In the following sections, we will describe these layers, mapping functions, and occurrences of these fatal pathologies. Summaries of observables and model parameters are shown in Table \ref{tab:param_summary} for easy reference.

@@

 \begin{figure}
 	\begin{center}
 		\includegraphics[width=\columnwidth]{tikz-figure0.pdf}
 	\end{center}
 	\caption{Probabilistic graphical model for our likelihood. Double-lined nodes represent observed variables, diamond nodes represent deterministic variables, single-lined ellipse nodes represent fit variables. The SN box represents observed variables and latent variables for each individual supernova, whilst the survey box represents survey-specific variables, which in general describe the supernova population for the survey and the systematics associated with it. Variables that appear outside both boxes represent top level model parameters. \added{To solidify that the underlying supernova population lives in $\lbrace M_B, x_1, c\rbrace$ space, we write out the stretch and colour populations along with $M_B$, however note here that those $x_1$ and $c$ values (denoted by an asterix) are identically equal to the $x_1$ and $c$ values in the latent node they are conditioned on, and in the code implementation, they are the same variable.}}
 	\label{fig:pgm}
 \end{figure}

@@

@@

\subsection{Supernova Selection Effects}
\label{app:selection2}
\added{
To turn the generalised equations from the previous sections into selection effects relevant for this model, need to highlight that $\theta$ represents only our top level parameters ($\Omega_m$, $w$, $\alpha$, $\beta$, etc), and that our parametrisation of true underlying values (for example $m_B$) takes the form of latent parameters. We thus continue from equation \eqref{eq:latent} and write out denominator $d$:
\begin{equation}
d = \iint  P(S|D, L, \theta) P(D | L, \theta) P(L | \theta)\, dD\, dL.
\end{equation}
Here we can now move from the generic labels $D$ and $L$ to something specific to our model. In this case, $D$ would represent our observables, $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{m}, \hat{p}\rbrace$, and $L$ our latent parameters $\lbrace m_B, x_1, c, z\rbrace.$ We assume that our selection effects can be sufficiently encapsulated by latent parameters. That is, we simplify $P(S|D,L, \theta) \rightarrow P(S|L)$. Specifically, we assume that our selection effects can be quantified using the true apparent magnitude $m_B$ and redshift $z$, and due to this we will not write out the other parameters in our model for simplicity. If we did, we would simply find them integrated to unity. Formally, we assume that $P(S|D, L, \theta) = P(S|z) P(S|m_B)$.}
Writing out the relevant observables $\hat{m}_B$, $\hat{z}$ and latent parameters $m_B$ and $z$, our denominator then becomes
\begin{equation}
d = \int P(S|z) P(S|m_B) P(\hat{z}|z) P(\hat{m}_B|m_B) P(z, m_B | \theta)\, d\hat{z} \, d\hat{m}_B \, dz \, dm_B.  \label{eq:sel} 
\end{equation}
Due to our assumed perfect measurement of redshift, $P(\hat{z}|z) = \delta(\hat{z} - z)$. $P(\hat{m}_B | m_B)$ is a Gaussian due to our Gaussian model of summary statistics, and can be analytically integrated out, collapsing the integral over $\hat{m}_B$. Finally, we can express $P(z, m_B | \theta)$ as  $P(m_B | z, \theta) P(z | \theta)$, where the first term requires us to calculate the magnitude distribution of our underlying population at a given redshift, and the second term is dependent on survey geometry and supernovae rates. We can thus state
\begin{equation}
d = \int \left[ \int P(S|m_B) P(m_B | z, \theta)\, dm_B \right] P(S|z)P(z|\theta)\, dz.
\end{equation}
By assuming that the distribution $P(S|z)P(z|\theta)$ is well sampled by the observed supernova redshifts, we can approximate the integral over redshift by evaluating
\begin{equation}
\int P(S|m_B) P(m_B | z, \theta)\, dm_B \label{eq:selint}
\end{equation}
for each supernova in the dataset -- i.e. Monte Carlo integration with assumed perfect importance sampling.

>>>>

At the most fundamental level, a supernova cosmology analysis is a mapping from an underlying population onto an observed population, where cosmological parameters are encoded directly in the mapping function. The difficulty arises both in adequately describing the biases in the mapping function, and in adding sufficient, physically motivated flexibility in both the observed and underlying populations whilst not adding \textit{too} much flexibility, such that model fitting becomes pathological due to increasing parameter degeneracies within the model.

\added{In the analysis of this article, the underlying model universe maps to the observed universe as sketched in the BHM of Figure \ref{fig:pgm}. The dependencies that between the model and observations can be tracked following the arrows of the BHM, and a summary of all the conditional probabilities can be found in \ref{sec:model_summary}.

In the following sections, we will describe the model parameters, the mapping functions that connect them to data, the effect of sample selection (in Equation \eqref{eq:main}), and the pathologies that can occur when evaluating the model. Summaries of observables and model parameters are shown in Table 1 for easy reference.}

@@

 \begin{figure}
 	\begin{center}
 		\includegraphics[width=\columnwidth]{tikz-figure0.pdf}
 	\end{center}
 	\caption{Probabilistic graphical model for our likelihood \added{and selection effects}. Double-lined nodes represent observed variables, diamond nodes represent deterministic variables, single-lined ellipse nodes represent fit variables. The SN box represents observed variables and latent variables for each individual supernova, whilst the survey box represents survey-specific variables, which in general describe the supernova population for the survey and the systematics associated with it. Variables that appear outside both boxes represent top level model parameters. \added{To solidify that the underlying supernova population lives in $\lbrace M_B, x_1, c\rbrace$ space, we write out the stretch and colour populations along with $M_B$, however note here that those $x_1$ and $c$ values (denoted by an asterix) are identically equal to the $x_1$ and $c$ values in the latent node they are conditioned on, and in the code implementation, they are the same variable.}}
 	\label{fig:pgm}
 \end{figure}

@@

\added{
\subsection{Model Summary}
\label{sec:model_summary}

Having laid out each individual aspect of the model, the relationships between variables and our treatment of uncertainty, here we summarise the relationships in our model likelihood mathematically. In this likelihood summary, we leave out sample selection for simplicity. Our likelihood is thus given as:

\begin{align}
&P({\rm data}|\theta) = \notag \\
&\int\, d\mu\, d\eta \ P( \hat{\eta}, \hat{\mathcal{C}}_\eta^{-1} | \eta, \delta \mathcal{Z}_i, \kappa_0, \kappa_1) P(\hat{z}|z) \notag \\
& P(\eta, \hat{m}, \hat{p} | \langle M_B \rangle, \langle x_1^i \rangle, \langle c^i \rangle, \mu, \delta(0), \delta(\infty), \sigma_{M_B}, \sigma_{x_1}, \sigma_c, \alpha, \beta) \notag \\
& P(\mu |z, \Omega_m, w) 
\end{align}
}

@@

\subsection{Supernova Selection Effects}
\label{app:selection2}
\added{
To turn the generalised equations from the previous sections into selection effects relevant for this model, need to highlight that $\theta$ represents only our top level parameters ($\Omega_m$, $w$, $\alpha$, $\beta$, etc), and that our parametrisation of true underlying values (for example $m_B$) takes the form of latent parameters. We thus continue from equation \eqref{eq:latent} and write out denominator $d$:
\begin{equation}
d = \iint  P(S|D, L, \theta) P(D | L, \theta) P(L | \theta)\, dD\, dL.
\end{equation}
Here we can now move from the generic labels $D$ and $L$ to something specific to our model. In this case, $D$ would represent our observables, $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{m}, \hat{p}\rbrace$, and $L$ our latent parameters $\lbrace m_B, x_1, c, z\rbrace.$ We assume that our selection effects can be sufficiently encapsulated by latent parameters. That is, we simplify $P(S|D,L, \theta) \rightarrow P(S|L)$. Specifically, we assume that our selection effects can be quantified using the true apparent magnitude $m_B$ and redshift $z$, and due to this we will not write out the other parameters in our model for simplicity. If we did, we would simply find them integrated to unity. Note that this assumption does not capture biases engendered by
Poisson noise fluctuations.  
We advocate that future analyses with
higher statistical precision do use a precisely determined $P(S|D)$.
Formally, we assume that $P(S|D, L, \theta) = P(S|z) P(S|m_B)$.}
Writing out the relevant observables $\hat{m}_B$, $\hat{z}$ and latent parameters $m_B$ and $z$, our denominator then becomes
\begin{equation}
d = \int P(S|z) P(S|m_B) P(\hat{z}|z) P(\hat{m}_B|m_B) P(z, m_B | \theta)\, d\hat{z} \, d\hat{m}_B \, dz \, dm_B.  \label{eq:sel} 
\end{equation}
Due to our assumed perfect measurement of redshift, $P(\hat{z}|z) = \delta(\hat{z} - z)$. $P(\hat{m}_B | m_B)$ is a Gaussian due to our Gaussian model of summary statistics, and can be analytically integrated out, collapsing the integral over $\hat{m}_B$. Finally, we can express $P(z, m_B | \theta)$ as  $P(m_B | z, \theta) P(z | \theta)$, where the first term requires us to calculate the magnitude distribution of our underlying population at a given redshift, and the second term is dependent on survey geometry and supernovae rates. We can thus state
\begin{equation}
d = \int \left[ \int P(S|m_B) P(m_B | z, \theta)\, dm_B \right] P(S|z)P(z|\theta)\, dz.
\end{equation}
By assuming that the distribution $P(S|z)P(z|\theta)$ is well sampled by the observed supernova redshifts, we can approximate the integral over redshift by evaluating
\begin{equation}
\int P(S|m_B) P(m_B | z, \theta)\, dm_B \label{eq:selint}
\end{equation}
for each supernova in the dataset -- i.e. Monte Carlo integration with assumed perfect importance sampling.
